% SIGPLAN format
\documentclass[oribibl]{llncs}

\usepackage[dvipsnames]{xcolor}

% AMS packages
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{mdwlist}

% Hyper links
\usepackage{url}
\usepackage{hyperref}
\hypersetup{
   colorlinks,
   citecolor=black,
   filecolor=black,
   linkcolor=black,
   urlcolor=black
}

% Miscellaneous
\usepackage{paralist}
\usepackage{graphicx}
\usepackage{float}
%\usepackage{titlesec}
%\usepackage[parfill]{parskip}
% \usepackage{balance} % Equalize column in the last page

% Revision tools
\usepackage{xspace}
\usepackage{comment}

\usepackage{algpseudocode}
\usepackage{longtable}

\newcommand\authornote[3]{\textcolor{#2}{#1: #3}}
\newcommand\bruno[1]{\authornote{Bruno}{red}{#1}}
\newcommand\ningning[1]{\authornote{Ningning}{blue}{#1}}
\renewcommand\arraystretch{1.1}
%\renewcommand\bruno[1]{}
%\renewcommand\ningning[1]{}

\let\oldpar\paragraph
\renewcommand{\paragraph}[1]{\oldpar{#1}\vspace{-10pt}}

% Literal programming support
\input{paper_utility.tex}
\usepackage[round, sort]{natbib}
\bibliographystyle {plainnat}

\begin{document}
\title{Sanitized Type Inference in Context\thanks{Research Paper. For formal review.}}

\author{Ningning Xie\thanks{Student paper.} \and Bruno C. d. S. Oliveira}
\institute{The University of Hong Kong}

\maketitle

\begin{abstract}
  Gundry et al. proposed type inference in context as a general foundation for
  unification/type inference algorithms. The key idea is based on the notion of
  information increase. Following this work, a more syntactic foundation for
  information increase is presented by Dunfield and Krishnaswami to deal with higher rank polymorphism.
  However, no existing work shows how it this idea can be extended naturally to deal
  with dependencies between existential variables in the presence of dependent types.
  In this paper, we propose a strategy called \textit{type
    sanitization} that helps resolve this problem
  in the framework of type inference in context. We show
  that type sanitization works on a unification algorithm for a dependently typed
  system with alpha-equality and first-order constraints. We then further illustrate
  this strategy to deal with polymorphic subtyping in a higher ranked
  polymorphic type system.
\end{abstract}

\input{sections/formal.tex}
\input{sections/lemmas.tex}

\setlength{\belowcaptionskip}{-1pt}
\input{sections/introduction.tex}
\input{sections/overview.tex}
\input{sections/dependent.tex}
\input{sections/higherrank.tex}
\input{sections/discussion.tex}
\input{sections/related_work.tex}
\input{sections/conclusion.tex}

% \input{sections/rules.tex}

\bibliography{sections/citation}

\appendix

\input{sections/complete_definitions.tex}
\input{sections/appendix.tex}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
