\section{Completeness and Soundness}

\subsection{Context Extension}

\begin{figure*}[h]
    \headercapm{\tctx \exto \ctxr}{Context Extension}

    \[\CEEmtpy \quad \CEVar\quad\CETypedVar\]
    \[\CETVar \quad\CEEVar \quad \CESolvedEVar\]
    \[\CESolve \quad\CEAdd \quad \CEAddSolved\]
    \caption{Context Extension.}
    \label{fig:ctx-extension}
\end{figure*}

\begin{figure*}[h]
    \headercapm{[\cctx]\tctx = \dctx}{Apply a complete context to a context}

    \begin{mathpar}
    \begin{tabular}{r c l l}
        $[\ctxinit]\ctxinit$   & = & $\ctxinit$    \\
        $[\cctx,x](\tctx,x)$ & = & $[\cctx]\tctx$  \\
        $[\cctx,\typA](\tctx,\typA)$ & = & $[\cctx]\tctx, \typA$  \\
        $[\cctx,x:\tau_1](\tctx,x:\tau_2)$ & = & $[\cctx]\tctx,x:[\cctx]\tau_1$ & if $[\cctx]\tau_1$=$[\cctx]\tau_2$ \\
        $[\cctx,\genA=\tau_1](\tctx,\genA)$ & = & $[\cctx]\tctx$ \\
        $[\cctx,\genA=\tau_1](\tctx,\genA=\tau_2)$ & = & $[\cctx]\tctx$ & if $[\cctx]\tau_1$=$[\cctx]\tau_2$ \\
        $[\cctx,\genA=\tau_1]\tctx$ & = & $[\cctx]\tctx$ & if $\genA \notin dom(\tctx)$ \\
    \end{tabular}
    \end{mathpar}
    \caption{Apply complete context.}
    \label{fig:apply-complete-ctx}
\end{figure*}

In order to prove completeness and soundness, we need a translation between declarative system and algorithmic system.

In Figure \ref{fig:algo-syntax}, we give the definition of complete context $\cctx$ below algorithmic context. The main difference between complete context and context is that a complete context has only solved existential variables. Since complete context has all existential variables solved, applying a complete context $\cctx$ to an algorithmic type $\tau$ will result in a type without existential variables $[\cctx]\tau$, which is well-formed under a declarative context.

We can extend the algorithmic output context to gain a complete context by $\tctx \exto \cctx$, which is shown in Figure \ref{fig:ctx-extension}. All original information will be preserved during context extension, including the existence of variables, their relative orders, type annotations, solutions for existential variables and all other information. In the meanwhile, information could be increased. Context extension could add more existential variables, or try to solve unsolved existential variables with proper types.

\rul{CE-Empty} states that empty context could be extended to empty context. \rul{CE-Var} preserves the variable. \rul{CE-TypedVar} is able to replace the type annotation with a contextually equivalent one. \rul{CE-LetVar} has two types that could be replaced. \rul{CE-EVar} and \rul{CE-SolvedEVar} keeps the existential variable. \rul{CE-SolvedEVar} solves existential variables with a type, requesting the type is well formed. \rul{CE-Add} and \rul{CE-AddSolved} adds one existential variable with unsolved and solved form respectively.

Obtaining a declarative context from an algorithmic system is similar to context subsitution, which is to apply the derived complete context $\cctx$ to algorithmic context $\tctx$ to get $[\cctx]\tctx$. Applying a complete context to a algorithmic context is defined in Figure \ref{fig:apply-complete-ctx}. Specifically, applying a complete context to itself will apply all the solved existential variables to types before throw away those existential variables. Notice this relationship $[\cctx]\tctx$ only exists when $\tctx \exto \cctx$. And we could have:

\begin{lemma}[Stability of Complete Contexts]
  \label{lemma:stability-of-complete-context}

  If $\tctx \exto \cctx$, then $[\cctx]\tctx = [\cctx]\cctx$.
\end{lemma}

\subsection{Soundness}

Having the notation of complete context, this section gives the lemma about the soundness of algorithmic system with respect to the declarative system, namely soundness of unification and soundness of typing.

\begin{theorem}[Soundness of Unification]

If $\tctx \byuni \tau_1 \lt \tau_2 \toctx$ where $[\tctx] \tau_1 = \tau_1$ and $[\tctx] \tau_2 = \tau_2$,
and $\ctxl \exto \cctx$,
then $[\cctx]\ctxl \bywf [\cctx]\tau_1$ and $[\cctx]\ctxl \bywf [\cctx]\tau_2$ and $[\cctx]\tau_1 = [\cctx]\tau_2$.
\end{theorem}

This theorem states that, if in algorithmic system, with context $\tctx$, unify $\tau_1$ and $\tau_2$ results in a output context $\ctxl$, which is extended to complete context $\cctx$, then we have the two types $[\cctx]\tau_1$, $[\cctx]\tau_2$ are well formed under declarative context $[\cctx]\ctxl$, and they are alpha equivalent.

\begin{theorem}[Soundness of Instantiation]

If $\tpreinst \sigma \lt \tau \toctx$ where $\tprewf \sigma$ and $\tprewf \tau$,
and $\ctxl \exto \cctx$,
then $[\cctx]\ctxl \byinst [\cctx]\sigma \lt [\cctx]\tau$.
\end{theorem}

If in algorithmic system, well-formed type $\sigma$ could be instantiated to another well-formed type $\tau$ with output context $\ctxl$, and $\ctxl$ extends to complete context $\Omega$, then in declarative system, the relationship of corresponding context and types also holds.

\begin{theorem}[Soundness of Generalization]

If $\tpreinf e:\tau \toctx $ and $\ctxl \bygen \tau \lt \sigma$,
and $\ctxl \exto \cctx$,
then $[\cctx]\ctxl \bygen [\cctx]e:[\cctx]\sigma$.
\end{theorem}

Generalization has different forms in two systems, so we need to an extra expression $e$. If in algorithmic system, expression $e$ has type $\tau$ and the latter could be generalized to type scheme $\sigma$, then in declarative system, the corresponding expression $e$ could have a correspondingly generalized type.

\begin{theorem}[Soundness of Typing]

Given $\ctxl \exto \cctx$
\begin{enumerate}
   \item If $\tpreinf e: \tau \toctx$ then $[\cctx]\ctxl \byinf [\cctx]e: [\cctx]\tau$
   \item If $\tprechk e: \tau \toctx$ then $[\cctx]\ctxl \bychk [\cctx]e: [\cctx]\tau$
   \item If $\tpreinf e_1:\tau_1 \toctx_1$ and $\ctxl_1 \byapp [\ctxl_1]\tau_1 ~ e_2 : \tau_2 \toctx$,
       then $[\cctx]\ctxl \byinf [\cctx]e_1 ~ [\cctx]e_2 : [\cctx]\tau_2$
\end{enumerate}
\end{theorem}

Soundness of typing has three forms corresponding to two modes in typing procedure and the application form. It states if in algorithmic system, with context $\tctx$, we could infer expression $e$ of type $\tau$, then in declarative context $[\cctx]\ctxl$, we could infer $e$ of type $[\cctx]\tau$. The statement is similar in rest cases.

\subsection{Completeness}

To prove the completeness of the algorithmic system with respect to the declarative system, we need to state that given a declarative typing judgement, we could construct an algorithmic one.

\begin{theorem}[Completeness of Unification]

If $\tctx \exto \cctx$ and $\tctx \bywf \tau_1$ and $\tctx \bywf \tau_2$ and $[\cctx]\tau_1 = [\cctx]\tau_2$, then there exist $\ctxl$ and $\cctx'$ such that $\ctxl \exto \cctx'$ and $\cctx \exto \cctx'$ and $\tctx \byuni [\tctx]\tau_1 \lt [\tctx]\tau_2 \toctx$.
\end{theorem}

\begin{theorem}[Completeness of Instantiation]

If $\tctx \exto \cctx$ and $[\cctx]\tctx \byinst [\cctx]\sigma \lt [\cctx]\tau$
then there exists $\ctxl$ and $\cctx'$
such that $\ctxl \exto \cctx'$ and $\cctx \exto \cctx'$ and $\tctx \byinst [\tctx]\sigma \lt [\tctx]\tau \toctx$.
\end{theorem}

\begin{theorem}[Completeness of Generalization]

If $\tctx \exto \cctx$ and $[\cctx] \tctx \bygen [\cctx]e \lt [\cctx]\sigma$,
then there exists $\ctxl$ and $\cctx'$ and $\tau$
such that $\ctxl \exto \cctx'$ and $\cctx \exto \cctx'$ and $\tctx \byinf e : \tau \toctx$ and $\ctxl \bygen \tau \lt \sigma$.
\end{theorem}

\begin{theorem}[Completeness of Typing]

Given $\tctx \exto \cctx$
\begin{enumerate}
    \item If $\tctx \bywf \tau$ and $[\cctx]\tctx \bychk [\cctx]e : [\cctx]\tau$\\
          then there exists $\ctxl$ and $\cctx'$\\
          such that $\ctxl \exto \cctx'$ and $\cctx \exto \cctx'$ and $\tctx \bychk e: [\tctx]\tau \toctx$.
    \item If $[\cctx]\tctx \byinf [\cctx]e : \tau$\\
          then there exists $\ctxl$ and $\cctx'$ and $\tau'$\\
          such that $\ctxl \exto \cctx'$ and $\cctx \exto \cctx'$ and $\tctx \byinf e:\tau' \toctx$ and $\tau = [\cctx']\tau'$.
    \item If $[\cctx]\tctx \byinf [\cctx]e_1 ~ [\cctx]e_2 :\tau_1$\\
          then there exists $\ctxl$ and $\ctxl_1$ and $\cctx'$ and $\tau_1'$ and $\tau_2$\\
          such that $\ctxl \exto \cctx'$ and $\cctx \exto \cctx'$ and $[\cctx']\ctxl_1 \byinf [\cctx]e_1 :[\cctx']\tau_2$ and $\tctx \byinf e_1 : \tau_2 \toctx_1$ and $\ctxl_1 \byapp [\ctxl_1]\tau_2 ~ e_2 :\tau_1' \toctx$ and $\tau_1 = [\cctx']\tau_1'$.
\end{enumerate}
\end{theorem}
